---
title: "Classification of Soccer Player Position by Numeric Metrics"
author: "Amal Babu, Ani Thakur"
date: "5/11/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r echo=FALSE, message=FALSE}
#install.packages("RSQLite")
library(RSQLite)
library(dplyr)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(car)
library(ggpubr)
library(cluster)
library(factoextra)
library(caret)
library(plotROC)
library(knitr)
```
## Introduction
As both of us are have played soccer for most of our lives, we thought it would be interesting to look at characteristics of professional soccer players around the world. We looked at 3 datasets from Kaggle. The first, called Soccer_Stats looks at numeric metrics in the 2015-2016 Premier League season for each FIFA player, such as overall rating, preferred foot, crossing, finishing, standing tackle, etc. We selected the data to only look at 12 key metrics, namely crossing', 'finishing', 'dribbling', 'sprint_speed', 'stamina', 'strength','aggression', 'vision', 'marking', 'standing_tackle', 'player_name', 'Position', 'Appearances' The second dataset called "player_pos" looks at variables such as team, position, appearances, goals, etc. for each professional player in the Premier League in the 2015-2016 season This dataset had 4179 observations that we selected on three variables, "player," "position," and "appearances." The third dataset looks at each player name and their FIFA ID number. This dataset allowed us to merge the two previous datasets upon the player name. To clean and tidy the data after merging we had to remove all null rows, remove duplicate rows for each player as FIFA constantly updates rankings to isolate one set of observations per player, and create a randomized smaller dataset to reduce the number of data points. Before beginning the analysis, some trends we expected were strong relationships between the finishing, dribbling, and sprint speed, as those players are primarily attackers and require those skills in abundance. On the defensive side we expected a strong relationship between a player's strength, aggression, and standing tackle, as those are important attributes for defenders. 

We also expected distinct clustering of players by position, as FIFA normally classifies player heavily based on what their position is. Some categories are more attacking based while others are defensive, and we expected that these variables will lead to significant variance between attackers and defenders. Finally, we expected that goalkeepers will be far away from all other players, as their skillset is entirely unique.

## Connect to Database
```{r}
connection <- dbConnect(RSQLite::SQLite(), dbname="database.sqlite")
```

## Gathering Datasets
```{r}
soccer_stats <- tbl(connection, 'Player_Attributes')

player_pos <- read.csv('Player Stats 2015-16.csv')%>%
  select(c('Player', 'Position', 'Appearances'))

player_id <- tbl(connection, 'Player')%>%
  select(c('player_name', 'player_fifa_api_id'))
```

## Merging Datasets by Player ID

```{r}
soccer <- left_join(soccer_stats, player_id, by = c('player_fifa_api_id'))
soccer <- left_join(soccer, player_pos, by=c('player_name'='Player'), copy=TRUE)
```

## Data Cleaning/Tidying
We selected 10 integer variables, as well as a players Position and Name which are chr. Variables. We converted all the integer variables to numerics, and deleted repeating entries for the same player, keeping only the first observation for each player. There are many observations because FIFA updates ratings multiple times throughout the year. Finally, to get rid of data points we randomnly sampled 20% of the data to keep a smaller dataset.
```{r}
# select important attributes
soccer <- soccer %>%
  select(c('crossing', 'finishing', 'dribbling', 'sprint_speed', 'stamina', 'strength',
           'aggression', 'vision', 'marking', 'standing_tackle', 'player_name', 'Position', 'Appearances'))
#save as data frame
soccer <- as.data.frame(soccer)


# convert integer rows to numeric, drop NA's
soccer <- soccer %>%
  mutate(across(.cols=where(is.integer), .fns = as.numeric))%>%
  drop_na()


# players have multiple entries b/c fifa updates rankings: keep only one ranking per player,
# the first ranking
soccer <- soccer %>%
  group_by(player_name)%>%
  filter(row_number() == 1)%>%
  select(-Appearances)

kable(head(soccer))


set.seed(322)
# create smaller dataset to reduce data points
small_soccer <- soccer %>% ungroup %>% sample_frac(0.2)

#check how many unique categories there are
unique(soccer$Position)
```
## Exploratory Data Analysis
We look at relationships between the main variables in our soccer dataset by looking at a correlation matrix as well as univariate/bivariate graphs. The visualizations show that player's "marking" ability is very highly correlated with a player's standing tackle ability. We also find high correlation between a player's "dribbling" and "crossing," "dribbling" and "finishing," and a player's "standing tackle" and "aggression" abilities. We find a low degree of correlation between "vision" and "marking" as well as "finishing" and "strength." The Pairs Panels further shows that  most variables follow a normal distribution,though skewed, except the "marking" and "standing tackle" variables, which follow a u-shaped distribution.

```{r}
# Correlation Matrix
soccer2 <- soccer %>%
  select_if(is.numeric) %>%
  ungroup()
soccer2 <- soccer2[-c(1)]           
cor(soccer2, use = "pairwise.complete.obs") %>%
  as.data.frame %>%
  rownames_to_column %>%
  pivot_longer(-1, names_to = "other_var", values_to = "correlation") %>%
  ggplot(aes(rowname, other_var, fill = correlation)) +
  geom_tile() +
  scale_fill_gradient2(low="red",mid="white",high="blue") +
  geom_text(aes(label = round(correlation,2)), color = "black", size = 4) +
  labs(title = "Correlation matrix for Soccer Player Characteristics", x = "variable 1", y = "variable 2")

## Univariate/Bivariate Graphs
# install.packages(psych)
library(psych)
pairs.panels(soccer, 
             method = "pearson", 
             hist.col = "magenta", 
             smooth = FALSE, density = FALSE, ellipses = FALSE)
```


## Setting up the Clustering

Initially we set up pam for the small_soccer dataset. However, we found that including Midfielders was a problem, as midfielders have attributes of both defenders and attackers, and are this hard to classify. Our final dataset was created by dropping all midfielders and classifying just forwards, defenders, and goalkeepers. You can see the intital clustering with k=4 below: notice how midfielders are scattered across 3 clusters.
```{r}
#PAM clustering
pam_results <- small_soccer%>%
  ungroup()%>%
  select_if(is.numeric)%>%
  scale()%>%
  pam(k=4)

fviz_cluster(pam_results, data=small_soccer, shape = 0) + 
  geom_point(aes(shape=small_soccer$Position))+
  guides(shape = guide_legend(title = 'shape'))

#get rid of midfielders: new dataset
small_soccer_no_mid <- small_soccer %>%
  filter(Position != 'Midfielder')
```

## Dimensionality Reduction
We inspected the principal components of the PCA algorithm and found that the "crossing," "dribbling," and "stamina" variables contributed the most variance to Dimension 1, which accounted for 52.6% of the explained variance. Dimension 2 accounted for 25.4% of the explained variance, and we found that the "marking," "standing tackle," and "finishing" variables contributed the most variance to this dimension. We retain PC 1 and PC2 to account for 78% of the total variance in the dataset. We identify low PC rates to have low levels of variance, and thus less distinguishing features across the clusters. We find the opposite to be true as well. 
```{r}
pca <- small_soccer_no_mid %>%
  select_if(is.numeric)%>%
  scale()%>%
  prcomp
get_pca_var(pca)$coord %>% as.data.frame

fviz_eig(pca,addlabels = TRUE)
head(small_soccer_no_mid)
```
## Clustering Visualizations
We performed PAM clustering on the positionality of the soccer player's, including whether they were a goalkeeper, defender, midfielder, or forward based on all the other attributes in the dataset. After initially running the algorithm we found that midfielders were hard to cluster because they have attributes of all players. The second time we ran the algorithm without the midfielders we found that the clusters were much easier to visualize. 
```{r}
small_soccer_no_mid <- small_soccer %>%
  filter(Position != 'Midfielder')
no_mid_pam <- small_soccer_no_mid %>%
  select_if(is.numeric)%>%
  scale()%>%
  pam(k=3)

cluster_column_pam <- small_soccer_no_mid %>%
  mutate(cluster = as.factor(no_mid_pam$clustering))

fviz_cluster(no_mid_pam, data=small_soccer_no_mid, shape = 0) + 
  geom_point(aes(shape=small_soccer_no_mid$Position))+
  guides(shape = guide_legend(title = 'shape'))
```
We also reported some summary statistics about the clusters; for example, for cluster 1 the highest mean was 'finishing' while the lowest mean was 'marking'. For cluster 2 the highest average was 'strength' and 'standing tackle' while the lowest was 'finishing'. Finally for cluster 3, the highest average was 'strenth' and the lowest was 'finishing'. While cluster 3 is harder to tell, it is easy to infer that cluster 1 is forwards and cluster 2 is defenders. We also found average width of clusters and the center point of each cluster

```{r}
kable(cluster_column_pam %>%
  group_by(cluster)%>%
  summarize_if(is.numeric, mean, na.rm=T))

no_mid_pam$silinfo$avg.width
no_mid_pam$medoids
```
## Classification and Cross-Validation
In order to do predict a binary variable, we must define a boolean variable. We decided to try to classify whether a player was a forward or not based on their stamina and finishing ratings. We appended a new new logical variable to our dataset representing whether that player is a forward or not.

```{r}
# add a variable which tells if someone is a forward or not
small_soccer_no_mid <- small_soccer_no_mid%>%
  mutate(is_forward = Position == 'Forward')

```

Next, we train a knn classifier. We split the data into train and test data and fit the knn model to the train data. We test on the test data, and record the accuracy, plot the ROC graph, and the boundary for the prediction line. We found that the model was able to classify players extremely accurately on just these two variables, with an ROC score of 0.967. Looking at the graph, we seem to be pretty accurate, with a small number of false positives and false negatives.
```{r warning=FALSE}
#KNN classifier

set.seed(322)

# Split data into train and test sets
train <- small_soccer_no_mid %>% sample_frac(0.8)
test <- small_soccer_no_mid %>% anti_join(train, by = "player_name")

# Fit the model on the train data
knn_fit <- knn3(factor(is_forward == 1, 
                       levels = c(TRUE,FALSE)) ~ stamina + finishing,
                data = train, 
                k = 5)

# Make a grid for the graph to layout the contour geom
grid <- data.frame(expand.grid(stamina = seq(min(small_soccer_no_mid$stamina),
                                            max(small_soccer_no_mid$stamina),
                                            length.out = 100),
                               finishing = seq(min(small_soccer_no_mid$finishing),
                                        max(small_soccer_no_mid$finishing),
                                        length.out = 100)))

# Use this grid to predict legendary status
grid %>% 
  mutate(p = predict(knn_fit, grid)[,1]) %>% 
  ggplot(aes(stamina, finishing)) + 
  # Only display data in the train set
  geom_point(data = train, 
             aes(stamina, finishing, color = as.factor(is_forward))) + 
  # Draw the decision boundary
  geom_contour(aes(z = p), breaks = 0.5) + 
  xlim(5, 100) + ylim(1,100) + 
  ggtitle("Train Set")

kNN_soccer <- test %>% 
  mutate(proportion = predict(knn_fit, test)[,1])


#visualize the ROC plot
ROCplot <- ggplot(kNN_soccer) + geom_roc(aes(d = is_forward, m = proportion), n.cuts = 0)
ROCplot

# calculate area under curve
calc_auc(ROCplot)$AUC
```

Finally, we preformed k-fold cross-validation with the same knn classifer, with 10 folds. Our average AUC was around 0.967 across all iterations. This is very good, almost a perfect classifier. We did not see many signs of overfitting, as there is very little variance in the AUC values
```{r warning=FALSE}
# KNN Classifier with cross-validation
# Set a seed to get reproducible results
set.seed(322)

k = 10

# Randomly order rows in the dataset
data <- small_soccer_no_mid[sample(nrow(small_soccer_no_mid)), ] 

# Create k folds from the dataset
folds <- cut(seq(1:nrow(data)), breaks = k, labels = FALSE) 

# Use a for loop to get diagnostics for each test set
diags_k <- NULL

for(i in 1:k){
  # Create training and test sets
  train <- data[folds != i, ] # all observations except in fold i
  test <- data[folds == i, ]  # observations in fold i
  
  # Train model on training set (all but fold i)
  fit <- glm(is_forward ~ stamina + finishing, 
             data = train, family = "binomial")
  
  # Test model on test set (fold i)
  df <- data.frame(
    prob = predict(fit, newdata = test, type = "response"),
    y = test$is_forward)
  
  knn_fit <- knn3(factor(is_forward == 1, 
                         levels = c(TRUE,FALSE)) ~ stamina + finishing,
                  data = train, 
                  k = 5)
  
  
  kNN_soccer <- test %>% 
    mutate(proportion = predict(knn_fit, test)[,1])
  
  
  # Consider the ROC curve for the test dataset
  ROCplot <- ggplot(kNN_soccer) + geom_roc(aes(d = is_forward, m = proportion), n.cuts = 0)
  
  
  # Get diagnostics for fold i (AUC)
  diags_k[i] <- calc_auc(ROCplot)$AUC
}

mean(diags_k)

```
